% ---- appendix.tex ----
% Data sources: /home/sharaths/projects/Prabh\=asa-integration/research/memory/journal.md (training config, cycle notes)
% /home/sharaths/projects/Prabh\=asa-integration/research/memory/findings.md (per-seed results F1/F2/F3)
% /home/sharaths/projects/Prabh\=asa-integration/site/src/data/results.json (official BLiMP, TextAvg scores)

\section{Training Configuration and Architecture}

All models are trained with a standard Transformer encoder with Rotary Position Embeddings (RoPE)~\cite{Su2021RoPE} using the HuggingFace Transformers library and custom training loops. The base configuration for both Strict and Strict-Small tracks is provided in \texttt{configs/phase2/} within the repository.

The Strict Track employs 14 layers, 768 hidden dimensions, and 12 attention heads with a feed-forward dimension of 3072. Positional encoding uses Rotary (RoPE). The tokenizer is a SentencePiece BPE with vocabulary size 20,000, shared with Strict-Small. The objective is Masked Language Modeling (MLM) only, without a causal language modeling (CLM) head. The masking schedule follows a cosine decay from 0.40 to 0.15 over training. The optimizer is Muon (Newton-Schulz orthogonalized momentum) with a learning rate of 1e-3 and linear warmup over 5\% of steps. Training uses a batch size of 256 sequences with maximum sequence length 192 tokens, running for 10 epochs and requiring approximately 13.7 hours on NVIDIA DGX Spark (Grace-Blackwell). The Strict Track employs a single seed (seed 0); additional seeds can be generated by sampling from BabyLM corpus subsets with different random states.

The Strict-Small Track contains 14 layers, 768 hidden dimensions, and 12 attention heads with the same feed-forward dimension of 3072 and RoPE positional encoding. It uses the same 20,000-token SentencePiece BPE vocabulary as the Strict track. The submission model employs hybrid MLM (50\%) plus CLM (50\%), while seed-wise comparisons use both pure-MLM and hybrid variants reported separately in the results. The masking schedule is identical (cosine decay from 0.40 to 0.15), as is the optimizer and learning rate (Muon, 1e-3, linear warmup). Batch size and sequence length remain 256 sequences and 192 tokens. Training runs for 10 epochs over the BabyLM Strict-Small corpus, with per-seed wall time approximately 82 minutes per epoch. Three independent seeds (random state 0, 1, 2) are used, with model checkpoints saved at epochs 5 and 10; the epoch-10 checkpoint is used for evaluation.

\section{Pāṇinian Mechanism Implementation}

The following optional components are controlled via command-line flags and are integrated into the training pipeline to investigate whether linguistic structure improves grammatical generalization.

The Vidyut N-hot Morpheme-Boundary Embeddings component, activated via \texttt{--use-nhot-embeddings}, detects morpheme boundaries using the Vidyut tokenizer, a lightweight Sanskrit morphological analyzer adapted for English character n-grams. Each token is embedded as a concatenation of a word-piece embedding (standard BPE) and a 10-dimensional one-hot morpheme-boundary indicator (for example, [1,0,0,...] for word-initial, [0,1,0,...] for word-medial, and so on). The morpheme embedding is projected to the model's hidden dimension via a learned linear layer and added to the standard BPE embedding before passing to the transformer encoder.

The Kāraka-Stratified Masking component, controlled by \texttt{--structured-masking}, stratifies token masking probability by crude syntactic role assignment. Tokens are classified into four bins: verb (kartā arguments), noun (karma arguments), adjective (visheshana modifiers), and separator (punctuation). The assignment uses a simple BPE-level lookup against a 200-word Sanskrit role lexicon weighted by Vedic corpus frequency. Each role is assigned a different masking probability; for example, kartā tokens are masked at 0.30, karma at 0.35, modifiers at 0.25, and separators at 0.10. These rates are configured to match the overall budget; if the global mask rate is 0.30, per-role rates are adjusted to achieve an effective mean of 0.30 via \texttt{--karaka-budget-match}. When enabled, this flag rescales per-role masking probabilities post-hoc to ensure the effective mean masking rate equals the scheduled global rate, ensuring fair comparison between structured and uniform baselines at identical effective mask budgets.

The Kāraka Auxiliary Objective, invoked via \texttt{--shabdabodha-aux LAMBDA}, adds a token-level kāraka-role prediction auxiliary task to the MLM loss when LAMBDA is greater than 0:
%
\begin{equation}
\mathcal{L}_{\text{total}} = \mathcal{L}_{\text{MLM}} + \lambda \cdot \mathcal{L}_{\text{aux}},
\end{equation}

where $\mathcal{L}_{\text{aux}}$ is the cross-entropy loss on 10-class role classification. The 10 classes are verb (kartā, karma, karan, etc.), noun, adjective, adverb, preposition, determiner, pronoun, punctuation, separator (implicit EOS), and unknown. Role labels are precomputed via spaCy dependency parsing on the training corpus. A \texttt{RoleStreamPacker} aligns these labels with the tokenizer's windowing scheme in lockstep, ensuring 1-to-1 alignment. The auxiliary head is a 2-layer MLP with hidden dimension 768 and output dimension 10, trained end-to-end with the main MLM head.

For specificity validation (F3), a control variant is trained with the same auxiliary objective but using random permutations of the role labels, preserving the label frequency distribution but destroying token-level alignment. This control tests whether the benefit is specific to linguistic role structure or attributable to generic multi-task regularization.

\section{Evaluation Methodology}

Models are evaluated on the official BLiMP test set comprising 67 paradigms across 12 grammatical phenomena. The primary metric is overall BLiMP, computed as mean accuracy across all 67 paradigms. A targeted subset of 20 paradigms from the agreement and argument-structure categories is analyzed separately as the focus for causal mechanism tests (F2 and F3). The BLiMP supplement comprises 14 additional paradigms testing rarer structures. Evaluation uses log-probability rather than rank: for each paradigm and example, the model assigns log-probabilities to the grammatical and ungrammatical variant, and accuracy is the proportion of examples where the grammatical variant receives higher log-probability.

Text Average is the macro-average over BLiMP, BLiMP Supplement, EWoK (Elements of World Knowledge)~\cite{Ivanova2024EWoK}, Entity Tracking, and COMPS (Conceptual Minimal Pair Sentences)~\cite{Misra2023COMPS} --- the official BabyLM ``text'' track aggregate. EWoK is a cognition-inspired world-knowledge benchmark; COMPS is a minimal-pair test of property knowledge; Entity Tracking measures zero-shot in-context coreference tracking over short passages~\cite{Warstadt2023}. For the Strict model, Text Average is 55.99; the Strict baseline achieves approximately 54.

\section{Empirical Results}

This section presents the supporting empirical detail behind the three findings reported in the main text, beginning with baseline comparisons and proceeding through the objective-scale, k\=araka-masking, and k\=araka-auxiliary analyses.

\subsection{Baseline Comparisons}

Models are also fine-tuned on GLUE (BoolQ, MultiRC, RTE, WSC, MRPC, QQP, MNLI, and QQPST) using 2-epoch fine-tuning with task-specific hyperparameters. Results are provided for reference but are not the focus of the main hypothesis tests. The Strict-Small submission model (prabhasa-b\_ss-0.1) achieves a GLUE average of 0.5807, computed as the mean of per-task best metrics. The BabyLM official baselines are Strict BLiMP 74.53 and Strict-Small BLiMP 65.08~\cite{Hu2024BabyLM}.

\subsection{Finding 1: Objective Scale-Dependence}

The comparison between pure MLM and hybrid MLM+CLM objectives demonstrates scale-dependent behavior. At 10M tokens (Strict-Small), pure MLM achieves a 3-seed mean of 63.58 ± 1.73 BLiMP (seeds 0/1/2 scoring 65.22/63.31/62.20; md5 checksums fa586488/15e33621/7a0bfa4a), while hybrid MLM+CLM achieves a 3-seed mean of 64.09 ± 0.26 BLiMP. The 95\% confidence intervals overlap substantially, indicating no significant difference at this scale. However, at 100M tokens (Strict track), pure MLM achieves 73.06 BLiMP, compared to the hybrid variant at 67.57, a difference of 5.49 percentage points. This finding indicates that the CLM head dilutes MLM quality in a scale-dependent manner, with the effect growing more pronounced as model scale increases. Table \ref{tab:f1-results} summarizes these results.

\begin{table}[h]
\centering
\small
\caption{Finding 1: Pure MLM vs Hybrid MLM+CLM at different scales.}
\label{tab:f1-results}
\begin{tabular}{lcccc}
\toprule
Objective & Scale & Mean BLiMP & SD & 95\% CI / Note \\
\midrule
Pure MLM & 10M & 63.58 & 1.73 & [61.85, 65.31] \\
Hybrid MLM+CLM & 10M & 64.09 & 0.26 & [63.83, 64.35] \\
Pure MLM & 100M & 73.06 & — & Single seed \\
Hybrid MLM+CLM & 100M & 67.57 & — & Single seed \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Finding 2: Kāraka Masking Causality}

Arm K (kāraka-stratified masking with budget matching) at 100M tokens achieves BLiMP 71.77 on the full test set and 82.03 on the targeted subset of 20 paradigms (agreement and argument-structure categories). Arm C (uniform masking with matched budget), also at 100M tokens, achieves BLiMP 70.49 and 81.93 on the targeted subset (both single seed). The causal contrast on the targeted subset is $\Delta_{K-C} = +0.10$ percentage points. A paired bootstrap analysis over the 20 paradigms with 1000 resamples yields a 95\% confidence interval of [-0.99, +1.20], and the $t$-test statistic is $t < 0.1$, indicating no significant difference. This result demonstrates that role-stratified masking probability, when properly budget-matched, provides no measurable causal improvement over uniform masking at this scale and configuration. Table \ref{tab:f2-results} presents this comparison.

\begin{table}[h]
\centering
\small
\caption{Finding 2: Kāraka masking (Arm K) versus uniform masking (Arm C) at 100M tokens.}
\label{tab:f2-results}
\begin{tabular}{lccc}
\toprule
Variant & Full BLiMP & Targeted Subset & Note \\
\midrule
Arm K (Kāraka-stratified) & 71.77 & 82.03 & Single seed, budget-matched \\
Arm C (Uniform) & 70.49 & 81.93 & Single seed, matched budget \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Finding 3: Kāraka Auxiliary Objective (Five-Seed Final)}

The kāraka auxiliary objective ($\lambda=1.0$) is evaluated against a baseline ($\lambda=0$) across five independent seeds (random states 0 through 4) at 10M tokens, with the corpus, the hybrid MLM objective, the masking, and the seed list held fixed; only the presence of the auxiliary objective varies. On the full 67-paradigm BLiMP suite, the auxiliary condition attains a 5-seed mean of 64.89 and the baseline 63.88, a difference of +1.01 percentage points. On the targeted 20-paradigm subset, the auxiliary condition attains 74.02 and the baseline 73.26. The single individually documented pair is seed 0, with auxiliary 74.97 versus baseline 72.32 on the targeted subset.

The pre-registered primary metric is the per-seed targeted-subset contrast ($\Delta_{\text{aux}-\text{base}}$). Across the five seeds the contrasts are +2.65, +0.61, +1.12, −0.22, and −0.34 percentage points, with a 5-seed mean of +0.76 (sd 1.21, se 0.54). A paired $t$-test yields $t=1.41$ (df=4) and a 95\% confidence interval of [−0.74, +2.27], which does not reach statistical significance. The contrast attenuates monotonically as seeds accumulate, from +2.65 at one seed to +1.46 at three seeds to +0.76 at five, indicating that the initial single-seed result reflected seed variance rather than a stable effect. The kāraka auxiliary objective therefore produces no statistically significant improvement over the baseline at five seeds and 10M token scale. Table \ref{tab:f3-results} summarizes these results.

\begin{table}[h]
\centering
\small
\caption{Finding 3: Kāraka auxiliary objective (treatment, $\lambda=1.0$) versus baseline ($\lambda=0$) across five seeds at 10M tokens.}
\label{tab:f3-results}
\begin{tabular}{lcccc}
\toprule
Condition & Full BLiMP & Targeted subset & Note \\
\midrule
Treatment (Aux, $\lambda=1.0$) & 64.89 & 74.02 & 5-seed mean \\
Baseline (No Aux, $\lambda=0$) & 63.88 & 73.26 & 5-seed mean \\
\midrule
Mean difference & $+1.01$ & $+0.76$ & sd 1.21, se 0.54 \\
95\% CI (targeted) & — & {$[-0.74, +2.27]$} & $t=1.41$, df=4, ns \\
\bottomrule
\end{tabular}
\end{table}

\section{Checkpoint Locations and Artifact Access}

The Strict-Small submission model is located at \texttt{qbz506/prabhasa-b\_ss-0.1} on HuggingFace, trained with hybrid MLM+CLM achieving 64.09 BLiMP (3-seed mean), with the repository including branches for per-seed checkpoints. The Strict track model at \texttt{qbz506/prabhasa-b\_s} on HuggingFace employs pure MLM and achieves 73.06 BLiMP as a single-seed run using the epoch 10 checkpoint. RQ-A and RQ-B checkpoints are available as revisions in the model cards, including all intermediate Arm K, Arm C, and auxiliary-objective variants (treatment and control). The grammar dataset is located at \texttt{qbz506/prabhasa-babylm-grammar} and contains morpheme inventories, role lexicons, and evaluation datasets.

\section{Code and Reproducibility}

The main repository is \texttt{github.com/SharathSPhD/prabhasa-babylm}. All code is version-controlled and tagged with the paper submission date (2026-06-15). The primary training entry point is the configuration-driven trainer in the repository, with mechanism flags documented inline and in the per-arm configuration files under \texttt{configs/phase2/}. Evaluation is handled by \texttt{scripts/official\_eval.py}, which reports BLiMP and Text Average metrics using the HuggingFace Evaluate library and official BabyLM benchmarking procedures. The analysis for F2 causal inference is in \texttt{scripts/analyze\_rqA.py} (per-paradigm bootstrap), and the F3 analysis is in \texttt{scripts/analyze\_rqB.py} (per-seed $t$-test). Adversarial review commentary on potential overclaims, alternative interpretations, and limitations is available in \texttt{docs/memory/tarka\_rqA\_rqB.md}.

\section{Ethical Considerations and Limitations}

The Prabhāsa models are trained on BabyLM corpora (English Wikipedia, Common Crawl, Books), which carry known biases and content restrictions. The models are not intended for production deployment and are released for research purposes only. The auxiliary mechanisms incorporate Sanskrit linguistic categories; users working in low-resource language settings should verify that role assignments are appropriate for their language family before adapting the code to other languages or domains. Model cards on HuggingFace include further details on intended uses, limitations, and potential biases.
