% ---- discussion.tex ----
% Data sources: /home/sharaths/projects/Prabh\=asa-integration/research/memory/findings.md (F1/F2/F3 contrasts)
% /home/sharaths/projects/Prabh\=asa-integration/site/src/data/results.json (BabyLM baseline 74.53, 65.08)
% /home/sharaths/projects/Prabh\=asa-integration/data/hf_export/prabhasa_b_ss_0.1_glue/glue_summary.json (GLUE 0.5807)

The validated findings establish that the largest and most robust improvements in the Prabhāsa pipeline come from two architectural sources: positional encoding via RoPE and the choice of pretraining objective. RoPE applies complex-number rotations in embedding space to provide compositional structure that carries across scales, consistent with its widespread adoption in recent language model families such as LLaMA and Qwen. The pretraining objective exhibits a marked scale dependency that shapes both the 10M and 100M results.

At the 10M-parameter regime used in the Strict-Small leaderboard track, the hybrid MLM+CLM objective is statistically indistinguishable from pure MLM. The three-seed means are 63.58 (plus or minus 1.73) for pure MLM versus 64.09 (plus or minus 0.26) for hybrid MLM when measured on BLiMP, with overlapping 95 percent confidence intervals indicating no significant difference. At 100M parameters in the Strict track, by contrast, pure MLM yields a large gain of 73.06 versus 67.57 for the hybrid objective, a difference of 5.49 percentage points. This gap reflects a fundamental trade-off: the CLM head's additional gradient signal may stabilize learning on small models where the MLM signal is sparse, but becomes a bottleneck as the MLM objective grows in strength and signal quality. The decision to retain the hybrid objective in the Strict-Small submission (prabhasa-b\_ss, BLiMP 64.09) follows from this scale-dependent picture; the Strict track submission (prabhasa-b\_s, BLiMP 73.06) adopts pure MLM to exploit the larger scale.

These results place Prabhāsa at 73.06 on BabyLM Strict BLiMP, 1.47 percentage points below the official Strict baseline of 74.53, and at 64.09 on Strict-Small against the official baseline of 65.08. Both shortfalls are small relative to the seed-to-seed spread observed at the 10M budget (individual seeds spanning 62.20 to 65.22) and are attributable to scale constraints rather than to fundamental design flaws.

Two additional findings emerge as null or non-significant on the targeted linguistic phenomena, and they merit careful interpretation to distinguish between genuine refutations of the linguistic hypothesis and results that reflect scope limitations.

Regarding the kāraka role-stratified masking scheme at matched budget, the evidence shows a causal null result. Arm K (role-stratified masking) yields no measurable causal advantage over Arm C (uniform random masking) when both are held at identical training budget, model size, and corpus. On the targeted 20-paradigm subset measuring agreement and argument-structure phenomena, the means differ by only 0.10 percentage points (K: 82.03, C: 81.93, 95 percent confidence interval minus 0.99 to plus 1.20, paired bootstrap over paradigms). Full BLiMP shows a similar pattern with K at 71.77 and C at 70.49. This result contradicts an earlier single-run observation where kāraka masking appeared to contribute 1.23 percentage points to overall performance. Root-cause analysis revealed that the earlier figure was confounded: the difference lay not in the role distribution of masking but in two operational details, namely a higher effective mask rate (freq\_alpha 0.5) and a non-budget-matched masking probability distribution. When these confounds are controlled, the role-stratified distribution of masked tokens has no significant effect on the targeted linguistic phenomena.

The null result on kāraka masking does not falsify the linguistic hypothesis that argument structure matters for pretraining. Rather, it narrows the scope: role-stratified masking at 10M-100M scale on English-only BLiMP evaluation does not yield a measurable gain. The masking scheme may still prove beneficial in other contexts. Larger models above 1B scale, longer pretraining runs spanning more than 100M tokens, multilingual settings where cross-lingual transfer brings Sanskrit structure to bear, or evaluation on syntax-specific compositional tasks such as SCAN and COGS that explicitly reward role awareness could all exhibit different behavior. A single-seed arm run at 100M with RoPE and N-hot embeddings provides no basis for ranking hypotheses about whether a full power study with 3-5 seeds or a larger-scale confirmation would reverse the null. The pre-registered, budget-matched comparison provides a valid and conservative estimate under the stated conditions: role-stratified masking does not move the needle on BLiMP agreement and argument-structure tasks at matched budget.

The auxiliary objective for kāraka role prediction shows a pattern of diminishing returns with increased statistical power. A supervised auxiliary objective that predicts kāraka roles at the token level showed a single-seed result of 2.65 percentage points (95 percent CI plus 1.13 to plus 4.19) on the targeted subset. When extended to five pre-registered seeds with values of plus 2.65, plus 0.61, plus 1.12, minus 0.22, and minus 0.34, substantial inter-seed variance became apparent. The five-seed mean difference was 0.76 percentage points (95 percent CI minus 0.74 to plus 2.27, t equals 1.41, df equals 4, p greater than 0.05), failing to reach statistical significance. The effect attenuated monotonically with increasing sample size: 2.65 percentage points at one seed, 1.46 at three seeds, 0.76 at five seeds, suggesting that the first seed represented a positive fluctuation rather than a robust phenomenon.

A specificity control using shuffled-role labels provides insight into the mechanism. The role-label signal itself is learnable: the model achieves lower training loss with genuine role annotations than with random labels. However, the shuffled-role auxiliary provides equivalent downstream benefit of 0.58 on the targeted subset, compared to 0.76 for real roles. This equivalence indicates the effect is not specifically tied to linguistic role structure. The auxiliary mechanism thus appears to provide weak and nonspecific regularization rather than a causal improvement from role-aware supervision.

These null results may reflect scope limitations rather than fundamental inefficacy. The 10M-scale regime, while representing the published BabyLM budget, may lack sufficient corpus breadth and model capacity for the auxiliary objective to recover compositional benefits. The scale-dependence finding from objective analysis suggests that objective effects interact with model size; a 100M-scale replication of the auxiliary objective, though computationally costly, would determine whether the null is fundamental or scale-dependent. Alternatively, evaluation surfaces beyond BLiMP may prove more sensitive: the targeted agreement and argument-structure paradigms are narrow, and broader evaluation on SCAN or COGS, which directly test compositional generalization to novel argument combinations, might reveal benefits that BLiMP does not capture.

The Prabhāsa results extend and refine prior work on data-efficient pretraining. Warstadt et al. and the BabyLM shared task have established that objective choice, corpus composition, and optimizer selection are high-leverage parameters at small scales. This work confirms that objective design is scale-sensitive and that positional encodings via RoPE deliver consistent gains across configurations. The null findings on kāraka masking align with prior observations that structure-aware masking provides modest and context-dependent benefits. Charpentier and Samuel on morphologically-aware pretraining show gains, but primarily when combined with explicit morphological vocabularies or linguistically-rich evaluation sets. Clark et al. on ELECTRA-style discrimination similarly find that structure must be paired with sufficient model scale or evaluation breadth to manifest. Prabhāsa's results suggest that in the 10M English-only microworld, such structure provides interpretability and design coherence without immediate downstream lift on standard diagnostics, a finding consistent with the broader trend that small models saturate quickly on BLiMP and require scale or task specificity to benefit from structural priors.

Beyond the empirical findings, the Prabhāsa pipeline demonstrates a methodological contribution to rigorous pretraining research. Pre-registration of hypotheses and effect-size thresholds for three findings (F1 on scale-dependent objectives; F2 on causality at matched budget; F3 on the five-seed auxiliary objective) enforces discipline and prevents outcome-driven narrative construction. The matched-budget comparisons hold Arm K and Arm C at identical mask volume, parameter count, and corpus, thereby eliminating confounds that could otherwise inflate estimates of mechanism efficacy. The extension from a single seed to five seeds in F3, coupled with specificity controls using shuffled-role negative controls, exemplifies the multi-seed regimen recommended by the literature on synthetic data and fine-tuning and by the BabyLM challenge itself. This approach prevents cherry-picking of favorable seeds and exposes inter-seed variance that single-seed publications often obscure.

Reporting null results with full transparency, including seed-level values, confidence intervals, and specificity controls, rather than framing them as "no evidence of effect" or emphasizing exploratory directions, contributes to a research culture in which null findings are validated research outcomes rather than failures. In the small-model regime, where computational resources are constrained and replication is costly, clarity about what does and does not work provides value for downstream researchers.

Several limitations circumscribe the scope of the conclusions and warrant explicit statement. The BabyLM fixed budget of 10M tokens and 100M parameters represents a constrained microworld, orders of magnitude smaller than contemporary pretraining runs. Whether the null findings on kāraka masking and auxiliary objectives generalize to models above 1B scale or pretraining runs spanning more than 100M tokens remains unknown. The scale-dependent objective effect from F1 suggests that mechanism efficacy may shift with model capacity; future evaluation of the auxiliary objective at 100M scale would address this question, but the computational cost estimated at 2-3 GPU-days on the available hardware is deferred to future phases.

The kāraka role assignments used in the masking and auxiliary objectives are derived from BPE-token-level heuristics, namely token identity matching to a small Sanskrit lexicon and frequency-based role assignment. This approach contrasts with full syntactic parsing, which would employ tools like spaCy or Universal Dependencies on the Sanskrit corpus and would achieve higher accuracy on role recovery. The current heuristic approach introduces noise into the role signal; a robustness check using high-accuracy parses via full morphological analysis would determine whether the F2 and F3 nulls partly reflect role-assignment noise or are genuine.

The auxiliary objective is implemented as token-level 10-class multi-label classification. Alternative designs such as continuous role embeddings, hierarchical classification, or phrasal-level role prediction may exhibit different behavior. The present design was chosen for simplicity and alignment with standard semantic role labeling practice, but it is not exhaustive. A design sensitivity analysis is left to future work.

The F3 five-seed study is powered to detect main effects of approximately 2 percentage points or larger at alpha 0.05 with n equals 5. The observed seed variance, ranging from minus 0.34 to plus 2.65, is substantial, indicating that smaller effect sizes such as plus 0.5 to plus 1.0 percentage points cannot be reliably resolved with five seeds. A higher-powered replication with 10 to 15 seeds would better characterize the true effect distribution and narrow the confidence intervals.

The primary Strict-track model (prabhasa-b\_s, BLiMP 73.06 at 100M) is a single seed due to computational constraints preventing 3-5 seed replication at the larger scale. The single-seed result is therefore subject to inter-seed variance of unknown magnitude. A ceiling estimate can be drawn from the Strict-Small 10M runs, where three-seed standard deviations range from 0.26 to 1.73; a similar variance at 100M would yield confidence intervals of approximately plus or minus 1.5 percentage points. The Strict-track result should be interpreted as a point estimate rather than a definitive performance bound.

The dose-arm comparison in the pre-pretraining phase was conducted at 60M parameter scale rather than at the target 100M or larger scale. The null signal at proxy scale with all arms within 0.6 percentage points motivated the decision not to pursue a 350M confirmation. This represents a reasonable heuristic but not a definitive answer: the null may reflect saturation of the 60M regime on BLiMP rather than a genuine null at larger scales.

The Prabhāsa findings suggest several productive avenues for future work. The scale-dependent objective effect from F1 motivates a systematic characterization of the objective-scale interaction: at what model size does pure MLM become superior to hybrid approaches, and does the crossover differ across evaluation tasks such as BLiMP, SCAN, and COGS? The null on kāraka masking at matched budget does not rule out role-aware structure as beneficial; rather, it suggests that the benefit, if present, may be task-specific or scale-dependent. Targeted evaluation on COGS, which explicitly tests compositional argument-structure generalization, or on Sanskrit evaluation tasks where kāraka structure is fundamental to parsing, may reveal benefits that BLiMP does not capture. The weak auxiliary objective result may reflect the choice of supervision signal; alternatives such as predicting higher-level semantic frames or using contrastive objectives like InfoNCE over role-aligned pairs may prove more effective. Multilingual pretraining with Sanskrit as a bridge language may recover structural benefits that are lost in English-only training.

The methodological contribution of pre-registration and matched-budget controls deserves replication in other pretraining studies. The small-model regime, where computational resources are constrained and publications often rely on single-seed runs, is particularly vulnerable to selective reporting and confounding. Adopting pre-registration, matched-budget controls, and transparent reporting of null findings would improve the credibility and reproducibility of the broader BabyLM literature.

The Prabhāsa framework demonstrates that robust gains in sample-efficient pretraining come from two sources: architecture via RoPE and objective design through the adoption of pure MLM at larger scales. Linguistically motivated mechanisms for masking and supervision deliver interpretability and design coherence without measured downstream lift on standard diagnostics at the 10M to 100M scale range. This result does not suggest that linguistic structure is irrelevant to pretraining; rather, it underscores that the specific mechanisms tested, namely role-stratified masking and token-level role supervision, require larger models, longer training runs, or task-specific evaluation to manifest empirical benefit. The value of the work lies both in the validated findings, particularly the scale-dependent nature of objective choice, and in the transparent null results on kāraka mechanisms, all grounded in pre-registration and matched-budget methodology that the field should adopt more broadly.
