% ---- introduction.tex ----
% Data sources: /home/sharaths/projects/Prabh\=asa-integration/research/memory/findings.md
% /home/sharaths/projects/Prabh\=asa-integration/site/src/data/results.json
% /home/sharaths/projects/Prabh\=asa-integration/.claude/worktrees/practical-mendel-bf73a8/paper/references_new.bib


The scaling of language model pretraining has driven rapid progress in natural language understanding and downstream task performance. Yet the relationship between data efficiency, inductive bias, and model behaviour remains underspecified, particularly at smaller scales. Small language models operating under strict computational budgets (10M to 150M parameters) offer a controlled experimental testbed for understanding which architectural and training choices yield measurable improvements, free from the confounds that emerge when training frontier-scale models. Understanding what matters at small scales is therefore both scientifically and practically important, as it constrains the design space for data-efficient pretraining and clarifies which inductive biases translate across scales.

One broad class of inductive biases derives from linguistic theory. Classical Pāṇinian grammar, developed over more than a millennium~\cite{Matilal2001}, encodes compositional structure at the morpheme level through morphological segmentation and inflectional rules, and at the argument level through systematic kāraka categories that denote semantic roles. These categories and their combinatorial rules are known to capture syntactic and semantic patterns in natural language. Whether embedding such structure as explicit inductive biases in the pretraining procedure improves data efficiency for small language models remains an open empirical question. The motivation is intuitive: if a linguistically grounded prior reduces the effective hypothesis space the model must search, it should require fewer training tokens to achieve a given level of generalisation.

This paper investigates this hypothesis through a controlled experimental study of Pāṇinian-motivated mechanisms within the BabyLM constrained-data setting~\cite{Warstadt2023}. The program, called Prabhāsa, comprises three research hypotheses that structure the empirical inquiry.

H1 (the primary hypothesis) asks whether morpheme-boundary structure, implemented as Vidyut N-hot embeddings~\cite{Vidyut} combined with role-aware masking (kāraka-stratified masking probability), measurably improves compositional generalisation on English syntax benchmarks (BLiMP agreement and argument-structure subsets) under matched training budgets and controlled comparisons. This hypothesis decomposes into two mechanistically distinct sub-questions that isolate different potential levers.

RQ-A (masking causality) asks whether at matched mask-rate budget, kāraka role-stratification of token masking carries causal efficacy on the targeted BLiMP paradigms. This question asks whether altering which tokens are masked, based on their morphological or semantic-role structure, affects downstream syntax performance.

RQ-B (auxiliary objective) asks whether a supervised auxiliary objective (token-level kāraka-role prediction from the contextualized embeddings of a pure-MLM model) improves BLiMP performance in cases where the masking lever from RQ-A was ineffective. This question asks whether explicit role supervision provides a signal the masking strategy alone does not.

H2 (synergy in scope) asks whether on the best H1-validated arm, a 6-phase Pañcāvayava fine-tuning scaffold reduces fallacious-inference rates and improves sample efficiency on reasoning-quality readouts versus a matched generic baseline.

H3 (epistemic constraint, out of scope) asks whether a GBNF schema coupled with a Z3 vyāpti verifier can enforce causal-validity constraints by construction.

The primary evaluation focuses on H1 with pre-registered thresholds such as at least +1.0 percentage point gain on targeted paradigm subsets, with significance assessed at at least 3 independent seeds using paired bootstrap resampling and Holm-Bonferroni family correction. Pre-registration serves a critical role: it prevents post-hoc rationalisation of weak or null findings and enforces statistical discipline before results are known.

This work contributes to the empirical understanding of linguistic inductive biases in small-model pretraining in several ways. First, RQ-A and RQ-B are tested independently using budget-matched arms to isolate the causal role of masking distribution from that of auxiliary supervision. This separation rules out confounds (such as differences in effective mask rate) that have plagued earlier single-arm comparisons and prevented clear mechanistic conclusions. Second, an orthogonal finding during the ablation revealed that the choice between pure masked-language modelling (MLM) and hybrid MLM plus causal-language-modelling (CLM) objectives is scale-dependent. At 10M tokens, the two objectives are statistically equivalent (pure-MLM 63.58 percent plus or minus 1.73 percent versus hybrid 64.09 percent plus or minus 0.26 percent on BLiMP, overlapping confidence intervals). At 100M tokens, pure-MLM substantially outperforms hybrid MLM plus CLM (73.06 percent versus 67.57 percent, a 5.49 percentage-point gain). This finding was validated independently and suggests that the CLM head's dilution of MLM training quality is a scale-dependent bottleneck in small-model pretraining. Third, all hypothesis tests employ at least 3 independent random seeds, paired bootstrap resampling within pre-registered target sets, and explicit pre-registration of outcome thresholds and inference procedures. A separate Tarka memo, representing the strongest objection to one's own finding, is prepared for each result to identify confounds and alternative explanations before closure. Fourth, code, data, model checkpoints, and per-seed results are released openly via HuggingFace model cards and the GitHub repository with full hyperparameter disclosure and configuration hashing, enabling replication and downstream use by the community.

To the best of the authors' knowledge, this is the first work to apply pre-registered, multi-seed, mechanistically-decomposed testing of Pāṇinian linguistic structure as an inductive bias for masked-language-model pretraining under strict data budgets. Prior work on linguistically motivated pretraining, including morpheme-aware objectives~\cite{Yaltik2020} and morphologically grounded word representations~\cite{Botha2014Morphology}, has tested linguistic priors in isolation or on larger models. Prabhāsa combines independent ablation of two mechanistically distinct levers (masking versus auxiliary objective), multi-seed validation, pre-registered significance thresholds, and explicit control for confounds. The distinctive contribution is therefore the mechanistic decomposition and statistical discipline, rather than any individual technique; together, these elements constitute a novel experimental design for isolating the causal efficacy of linguistically motivated structure in small-model pretraining.

The empirical findings from H1 indicate that architectural choices (RoPE positional encoding~\cite{Su2021RoPE}) and objective design (pure MLM without auxiliary CLM) are the primary levers for improving compositional generalisation at this scale (10M to 100M tokens). The Pāṇinian mechanisms, including masking stratification and auxiliary kāraka-role prediction, show null or marginal effects when tested with appropriate controls. Specifically, RQ-A (masking causality) yields a pre-registered null finding at 100M (Arm K versus Arm C: delta K minus C equals +0.10 percentage points, 95 percent confidence interval [negative 0.99, +1.20], non-significant). RQ-B (auxiliary objective) exhibits a weak, non-significant effect at 10M when tested across 5 seeds (delta Aux minus Base equals +0.76 percentage points, 95 percent confidence interval [negative 0.74, +2.27], t equals 1.41, df equals 4, not significant), with the single-seed positive result shrinking after adding more seeds, indicating seed variance rather than robust signal.

This outcome is scientifically valuable as it constrains the scope of claims about structure-driven pretraining, identifies the true levers (architecture and objective), and provides a rigorous baseline for null findings in the community. A well-documented null is a valid and necessary contribution, particularly when it overturns optimistic prior expectations and is grounded in pre-registered, multi-seed validation.

The remainder of this paper is structured as follows. Section~\ref{sec:related} reviews related work on inductive biases in pretraining, data-efficient language model training, and computational treatments of Pāṇinian grammar and Navya-Nyāya. Section~\ref{sec:data} describes the corpora, the synthetic Pāṇinian pre-pretraining dose, and the four dose arms. Section~\ref{sec:methods} introduces the model architecture and the Pāṇinian mechanisms, and states the research questions and their pre-registered thresholds. Section~\ref{sec:training} describes the training stages, the fine-tuning procedure, and the evaluation protocol. Section~\ref{sec:results} reports results for each research question with effect estimates and confidence intervals. Section~\ref{sec:discussion} discusses the interpretation of the findings, their implications for small-model pretraining, and threats to validity. Section~\ref{sec:conclusion} concludes and outlines future work. Supplementary configuration details, per-seed tables, and reproducibility information are provided in Appendix~\ref{app:repro}.
