The scaling of language model pretraining has driven rapid progress in natural language understanding, yet the relationship between data efficiency, inductive bias, and model behaviour remains underspecified at smaller scales. Small language models under strict computational budgets offer a controlled testbed for understanding which architectural and training choices yield measurable improvements, unclouded by the confounds that emerge at frontier scale. This work investigates whether linguistic structure from classical Pāṇinian grammar, developed over more than a millennium~\cite{Matilal2001}, operationalized through morpheme-boundary embeddings and role-aware masking, improves data-efficient pretraining of small models within the BabyLM constrained-data setting~\cite{Warstadt2023}.

Pāṇinian grammar encodes compositional structure at the morpheme level through inflectional rules and at the argument level through kāraka categories that denote semantic roles. The hypothesis is intuitive: if a linguistically grounded prior reduces the effective hypothesis space, the model should require fewer tokens to achieve given generalization. Prabhāsa investigates this through three research hypotheses. RQ-A (masking causality) tests whether kāraka role-stratified token masking, at matched mask-rate budget, carries causal efficacy on targeted BLiMP paradigms, operationalized via morpheme-boundary N-hot embeddings~\cite{Vidyut}. RQ-B (auxiliary objective) tests whether supervised auxiliary prediction of kāraka roles improves BLiMP performance when masking alone is ineffective. H2 (synergy, in scope) asks whether a Pañcāvayava fine-tuning scaffold improves reasoning quality on the best validated arm. H3 (epistemic constraint, out of scope) concerns constructive validity enforcement via Z3 verifiers. The primary evaluation focuses on H1 with pre-registered thresholds of at least plus 1.0 percentage point gain on targeted subsets, assessed at least 3 independent seeds using paired bootstrap resampling and Holm-Bonferroni correction.

This work contributes through mechanistic decomposition and statistical discipline. First, RQ-A and RQ-B are isolated using budget-matched arms to rule out confounds such as mask-rate variation that plagued prior comparisons of linguistically motivated masking~\cite{Yaltik2020, Botha2014Morphology}. Second, an incidental finding revealed that the choice between pure masked-language modelling (MLM) and hybrid MLM plus causal-language-modelling (CLM) is scale-dependent: at 10M tokens the two objectives show neutral difference (pure-MLM 63.58 plus or minus 1.73 percent versus hybrid 64.09 plus or minus 0.26 percent, overlapping confidence intervals), while at 100M tokens pure-MLM substantially outperforms hybrid MLM plus CLM (73.06 percent versus 67.57 percent, a 5.49 percentage-point gain). This suggests the CLM head becomes a scale-dependent bottleneck to MLM signal quality. Third, all hypothesis tests employ at least 3 independent random seeds, paired bootstrap resampling, and explicit pre-registration of thresholds. Fourth, code, data, and model checkpoints are released openly with full hyperparameter disclosure (see Appendix~\ref{app:repro}).

The empirical findings show that architecture (RoPE positional encoding~\cite{Su2021RoPE}) and objective design (pure MLM at scale) are the primary levers for improving BabyLM performance at 10M-100M parameters. The Pāṇinian mechanisms exhibit null or marginal effects under proper controls. Specifically, RQ-A yields a pre-registered null at 100M (Arm K versus Arm C, delta equals plus 0.10 percentage points, 95 percent confidence interval [negative 0.99, plus 1.20], non-significant). RQ-B exhibits a weak, non-significant effect at 10M across 5 seeds (delta equals plus 0.76 percentage points, 95 percent confidence interval [negative 0.74, plus 2.27], t equals 1.41, df equals 4, not significant). A well-documented null is a valid scientific contribution that constrains claims about structure-driven pretraining and identifies the true levers. The Strict-track submission (prabhasa-b\_s, 100M pure-MLM) achieves 73.06 BLiMP versus the Strict baseline of 74.53, and the Strict-Small submission (prabhasa-b\_ss-0.1, 10M hybrid) achieves 64.09 BLiMP versus the Strict-Small baseline of 65.08.

The remainder of this paper is organized as follows. Section~\ref{sec:related} reviews related work on inductive biases in pretraining, data-efficient training, and computational treatments of Pāṇinian grammar. Section~\ref{sec:methods} introduces the model architecture, the Pāṇinian mechanisms, and the training and evaluation protocol, and states the research questions with pre-registered thresholds. Section~\ref{sec:results} reports results with effect estimates and confidence intervals. Section~\ref{sec:discussion} interprets the findings and discusses implications and threats to validity. Appendix~\ref{app:repro} provides configuration details, per-seed tables, and reproducibility information.
